Eleven years of student replication projects provide evidence on the correlates of replicability in psychology

### 2023-05-03 paper/manuscript.tex
flial at intro

\begin{abstract}
Cumulative scientific progress requires empirical results that are robust enough to support theory construction and future experimental extension. Yet in psychology, some prominent individual findings are have repeatedly failed to replicate, and large-scale investigations suggest that the overall replication rate is around half.
What predicts which studies will replicate? The identification of features that predict replication is limited by the common dataset of only 170 studies from large-scale investigations that are repeatedly analysed.
We introduce a new sample of 176 replications conducted by students as part of a graduate-level experimental methods course, and we examine how often students were able replicate their chosen studies' results closely enough that they would be ready to build on it in future work. Using this as a metric of replicability, 49\% of the studies in our sample replicate, with studies with larger original effect sizes and within-participants designs being more likely to replicate.
Consistent with prior reports, our replication results indicate that the robustness of the psychology literature is low enough to limit progress by student investigators.

@@

Cumulative science requires foundation empirical results that are robust enough to support theory construction and future experimental extension. If scientists attempt to build on top of results that cannot be achieved consistently, they waste resources looking for results where there may be none. This is frustrating for individual scientists and damaging to the enterprise of science overall.

Scientists, policymakers, and the public often treat publication in a scientific journal as a signal that a result is (likely to be) true. This trust in published findings has led scientists to build on findings without verifying them first, but failures were often explained away and rarely published, leaving the original results uncontested in the literature, while failures to achieve the same outcomes accumulated in private. FIX PRIOR SENTENCE

During the replication crisis, publicized failures to replicate published findings called into question the tacit assumption that most of the literature was robust and would replicate. Certain prominent findings failed to replicate in multi-site replication attempts (ex. terror management theory, \protect\hyperlink{ref-klein2022}{Klein et al. 2022}, ego-depletion, \protect\hyperlink{ref-hagger2016}{Hagger et al. 2016}), and a large-scale replication project pegged the replication rate for findings in top-tier psychology journals at around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Open Science Consortium 2015}).

These failures to replicate raised the question: How replicable is psychology as a whole? Only three large-scale investigations address this question by sampling subsets of psychology findings. Reproducibility Project: Psychology, mentioned above, sampled roughly 100 studies from articles published in three top psychology journals in 2008 and found a overall replication rate of around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Open Science Consortium 2015}). ManyLabs investigated heterogeneity using short target studies that compared between two conditions each, but these studies are not representative of the psychology literature as a whole. Across Many Labs 1-3, 29 of 51 target effects (57\%) replicated (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) replicated all 21 behavioral social science studies from Nature and Science from 2010-2015 that were feasible; the replication rate was around 60\%. Conducting large numbers of studies to assess replication rates is expensive and arduous which may explain why more large-scale investigations don't exist. These roughly 170 replications from these three investigations are the primary empirical results on which discussions of replication rates in psychology are based.

TRANSITION A replication rate around 50\% runs counter to the prior assumption that most findings in the literature were robust enough to build upon without verification, and it raises the question: What makes the replication rates in psychology so low?

Experimental approaches attempt to causally intervene on the low replication rate, by changing certain factors that may affect whether studies replicate. Protzko et al. (\protect\hyperlink{ref-protzko2020}{2020}) showed, across 16 studies, that better methodological practices, such as transparency, large sample sizes, and confirmatory tests, led to replication rates that matched theoretical expectations based on the effect sizes and sample sizes, with replication effect sizes comparable with the original. Many Labs 5 examined how adding expert advice to a replication process might increase the replication rate, and found that it mainly did not for the 10 studies they looked at (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). These types of experiments are valuable for testing potential causes of non-replication, but they don't scale well due to expense. Additionally, not all potential influences on non-replication are experimentally manipulable.

Correlational approaches looking for predictors of replicability are more popular. Prediction markets and elicitations have established that people can predict what studies will replicate above chance (\protect\hyperlink{ref-dreber2015}{Dreber et al. 2015}, \protect\hyperlink{ref-camerer2018}{Camerer et al. 2018}, \protect\hyperlink{ref-forsell2019}{Forsell et al. 2019}, \protect\hyperlink{ref-hoogeveen2019}{Hoogeveen et al. 2019}), but have not identified concrete predictors that differentiate replications from non-replications. Using machine learning approaches, Altmejd et al. (\protect\hyperlink{ref-altmejd2019}{2019}) examined statistical and demographic features of studies and identified larger sample sizes, larger effect sizes, and simple effects (as opposed to interaction terms) as predictive of replication. Open Science Consortium (\protect\hyperlink{ref-openscienceconsortium2015}{2015}) looked at correlates of replicability in the RP:P sample and found that studies in cognitive psychology (as opposed to social psychology) and studies with larger effect sizes and smaller p-values were more likely to replicate. One potential set of correlates to replicabilty that has not been thoroughly examined are experimental design features, such as between or within subjects designs and repeated measures.

TRANSITION / HOW TO FIT THIS PART IN Correlational approaches depend on data from replications, generally drawing heavily from the same small set of data points. In particular, the RP:P dataset itself is much discussed and reanalyzed (\protect\hyperlink{ref-anderson2016}{Anderson et al. 2016}, \protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}) to the point that much of what we think we know about replicability may be overfit to the 100 studies included in in RP:P.

TRANSITION / DELETE ? Many of the discussions and reanalyses have raised questions about what the right metrics for measuring replicability are and what standards are reasonable to expect. Some of this contention is based around notions that failed replications are referenda on the truth of the original finding.

TRANSITION! Prior approaches to replicability have focused on interpreting results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (\protect\hyperlink{ref-vanbavel2016}{Van Bavel et al. 2016}) or even time period (\protect\hyperlink{ref-ramscar}{Ramscar et al. n.d.}).

Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes for prior work typically do not reflect an entirely random sample from the literature; instead they sample from specific journals where results may be of more interest and adjust the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of ``truth'' in the literature as a whole. Sampling truly at random from the literature may not even be desirable, as arguably a literature will succeed if useful discoveries come out of it, not if random findings are true (\protect\hyperlink{ref-wilson2020}{Wilson et al. 2020}). The importance of a study being replicable is not uniformly distributed across the literature.

In contrast to prior approaches, we have explicitly pursued a different estimand: the probability that a researcher, on selecting a finding of interest from the literature, can successfully achieve a result satisfactorily close enough to the original that they can build on it in their own work, with all the necessary compromises to the methods and sample of the original that may be required by the constraints of the situation.

NEED PUNCHY TRANSITION TO DETAILS Rather than sampling at random from some parts of the literature; our sample of studies is selected based on what studies students were interested in and wanting to replicate, with some filtering for feasibility; this sampling reflects how scientists choose what studies to build on: those that are interesting and relatively doable given methodological and budgetary constraints.

@@

All our replications were conducted under short time scales and relatively small budgets, which parallels the constraints many scientists are under when starting new projects. From the point of view of cumulative science, researchers want to know if they can get paradigms to work under their real-world constraints. The same issue that may cause a study not to replicate under constrained circumstances (despite hypothetically replicating under more favorable circumstances such as expert administration or larger budgets) will also plague attempts to build off those studies in constrained circumstances. Thus, we believe it is relevant to estimate replicability under the limitations of real world resource and expertise constraints.

@@

\hypertarget{discussioni}{%
\section{Discussioni}\label{discussioni}}

MAY NEED SMOOTHER TRANSITIONS THROUGHOUT
Prior large scale replications of psychology findings have estimated the replication rate at somewhere around one half, which intuitively feels low. However, limited numbers of replications and limited research into specific predictors of replication failure mean that the reasons for non-replications are not well understood.

@@

Our metric of replicability was operationalized around the idea of whether a study could support cumulative science, so we considered whether results similar to the original could be obtained under the resource, expertise, and time constraints on early stage graduate students. We think this is a valuable estimand that is consistent with scientists' reasons for valuing replicability, but it is not the only estimand. Other replication projects have estimated replicabilty in other circumstances; for instance, Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) looks at how likely studies are to replicate when performed by an expert with a large budget, and Ebersole et al. (\protect\hyperlink{ref-ebersole2020}{2020}) asks how likely certain studies are to replicate when performed with extensive feedback from the original authors.

Replication results should be interpreted in light of their methods and estimand. We do not interpret our non-replications as indicating the original results were false positives (presumably some were and some weren't). There are many possible reasons for the non-replications in this sample. In some cases, the problem may be with the replication, such as too few participants, many exclusions for failed attention checks, or participants speeding through the study. When these issues are diagnosed, they suggest possible ways to ``rescue'' the replication by increasing the sample or changing the interface, without altering the underlying experiment. In other cases, there were a priori reasons to distrust the original study, such as exclusion criteria that seemed post-hoc or high-order interaction terms with a small sample. That said, not all scientists recognize the same factors as potential indications of low power or questionable research practices; students conducting these replications generally expected them to succeed. In many non-replication cases, it was unclear why the results failed to replicate.

We also caution the interpretation of our predictor variables. Some of the correlates are most easily interpreted as being about the original study, and others reflect the closeness of the replication to the original. This was a correlational design, so we cannot attribute causality to any of these factors. For instance, while within-subjects designs are more likely to replicate than between-subjects designs, this could be related to power, or the types of experiments that tend to be run in each design. Given the predictive value, slightly more skepticism and critical reading of between-subjects designs may be warranted, but this correlation, by itself, does not mean scientists should prefer to run within-subjects designs.

Our results are limited by the number and quality of the studies we included. These studies are not necessarily representative of the studies of interest to psychologists as a whole. The replication studies were not designed from the start with this analysis in mind, so analysis is limited by the choices and reporting used at the time of replications. We reiterate that given our sample, we are able to estimate the correlates of studies replicating when the replications are conducted online, by graduate students with limited time and budget. This estimand is not the same as whether they would replicate when done by large labs with access to more money and experts. This estimand does not represent whether or not the original study was ``true''.

>>>>

\begin{abstract}
Cumulative scientific progress requires empirical results that are robust enough to support theory construction and future extensions.
Yet in psychology, some prominent individual findings have failed to replicate, and large-scale investigations suggest that failures to replicate are widespread.
What predicts which studies will replicate?
The identification of features that predict replication success is limited by data, however: most studies use a common dataset of only \textasciitilde170 studies, which vary in how they define replication.
We introduce a new dataset of 176 replications conducted by students as part of a graduate-level experimental methods course.
We examine how often students were able replicate their chosen studies' results closely enough that they would be ready to build on it in future work.
Using this as a metric of replicability, 49\% of the studies in our sample replicate, with studies with larger original effect sizes and within-participants designs being more likely to replicate.
Consistent with prior reports, our results indicate that the robustness of the psychology literature is low enough to limit progress by student investigators.

@@

Cumulative science requires foundational empirical results that are robust enough to support theory construction and future experimental extension. Scientists often treat publication in a scientific journal as a signal that a result was (likely to be) true, which leads scientists to build on findings without verifying them first. The replication crisis called into question the tacit assumption that most of the literature was robust and would replicate. Certain prominent findings failed to replicate in multi-site replication attempts (ex. terror management theory, \protect\hyperlink{ref-klein2022}{Klein et al. 2022}, ego-depletion, \protect\hyperlink{ref-hagger2016}{Hagger et al. 2016}), and a large-scale replication project pegged the replication rate for findings in top-tier psychology journals at around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Open Science Consortium 2015}). When the trust in the literature mismatches the robustness results, many scientists can waste time and resources becoming frustrated pursuing extensions to a popular, but unreliable result. This seems to be the current situation of the literature, which leads to question: what do we do to minimize the waste?

From a policy perspective, we need to understand the landscape of the problem before we can have research-informed science policy interventions. Thus, we might want to know 1) what quantity to measure if we want to capture frustration and waste due to non-cumulative findings and 2) from a descriptive perspective, what types of studies are most at risk for this issue. Such measurements have been the focus of much prior work, but there is limited consensus on what is being estimated.

Three large-scale investigations have addressed replication rates in psychology. Reproducibility Project: Psychology, mentioned above, sampled roughly 100 studies from articles published in three top psychology journals in 2008 and they found an overall replication rate of around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Open Science Consortium 2015}). ManyLabs investigated heterogeneity using short target studies that compared between two conditions each. These study designs were not representative of the psychology literature as a whole, and due to the heterogeneity goal, they had large overall samples across multiple sites. Across Many Labs 1-3, in these high-powered replications, only 29 of 51 target effects (57\%) replicated (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) replicated all 21 behavioral social science studies from Nature and Science from 2010-2015 that were feasible; they consulted original authors and had high power to detect effects smaller than those reported. In this well-resourced environment with expert input, the replication rate was around 60\%. These studies all measured replication rates, but the details of their methods and outcome measures varied.

While their sampling and methods varied, these previous approaches to replicability have focused on interpreting their results in terms of a potentially problematic estimand: the probability of a finding in the literature being somehow truly replicable. Critics have pointed out that ``true'' replicability may not be possible to estimate outside of a specific sample (\protect\hyperlink{ref-vanbavel2016}{Van Bavel et al. 2016}) or even time period (\protect\hyperlink{ref-ramscar}{Ramscar et al. n.d.}). Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes did not reflect an entirely random sample from the literature; instead replication projects sampled from specific journals where results may be of more interest and adjusted the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of ``truth'' in the literature as a whole. Sampling truly at random from the literature may not even be desirable, as arguably a literature will succeed if useful discoveries come out of it, not if random findings are true (\protect\hyperlink{ref-wilson2020}{Wilson et al. 2020}). The importance of a study being replicable is not uniformly distributed across the literature.

Rather than aim for some measure of ``true'' replicability, it makes more sense to contextualize replication efforts based on their methods and outcome measures. With this lens, we could say that RP:P is looking at how well findings in well-known journals can be replicated in typical circumstances. Many Labs looked at which well-known two conditions studies replicate with huge samples, and Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) was looking at what replicates when conducted in a highly-resourced environment with expert involvement. All of these are pontentially desirable estimands depending on what the specific circumstances and questions of interest are.

From the point of view of enabling cumulative science, an estimand of interest is how well work can be replicated by students, as students conduct most scientific work (\protect\hyperlink{ref-frank2022}{\textbf{frank2022?}}). In our current study, we use student success at replicating an effect as our estimand. In particular, our estimand is the probability that a researcher, on selecting a finding of interest from the literature, can successfully achieve a result satisfactorily close enough to the original that they can build on it in their own work, with all the necessary compromises to the methods and sample of the original that may be required by the constraints of the situation.

Our replication approach reflects this estimand. Our sample of studies is selected based on what studies students were interested in and wanting to replicate, with some filtering for feasibility; this sampling reflects how scientists choose what studies to build on: those that are interesting and relatively doable given methodological and budgetary constraints.

@@

All our replications were conducted under short time scales and relatively small budgets, paralleling the constraints many scientists are under when starting new projects. From the point of view of cumulative science, researchers want to know if they can get paradigms to work under their real-world constraints.

Despite variation in how large-scale replication studies have defined replication successes, there is work suggesting some consistency in the predictors of replication success. Knowing these predictors is important for appropriately calibrating how we read the literature and for eventually identifying targets for interventions.

Prediction markets and elicitations have established that people can predict what studies will replicate above chance (\protect\hyperlink{ref-dreber2015}{Dreber et al. 2015}, \protect\hyperlink{ref-camerer2018}{Camerer et al. 2018}, \protect\hyperlink{ref-forsell2019}{Forsell et al. 2019}, \protect\hyperlink{ref-hoogeveen2019}{Hoogeveen et al. 2019}), but have not identified concrete predictors that differentiate replications from non-replications. Using machine learning approaches, Altmejd et al. (\protect\hyperlink{ref-altmejd2019}{2019}) examined statistical and demographic features of studies and identified larger sample sizes, larger effect sizes, and simple effects (as opposed to interaction terms) as predictive of replication. Open Science Consortium (\protect\hyperlink{ref-openscienceconsortium2015}{2015}) looked at correlates of replicability in the RP:P sample and found that studies in cognitive psychology (as opposed to social psychology) and studies with larger effect sizes and smaller p-values were more likely to replicate. One potential set of correlates to replicabilty that has not been thoroughly examined are experimental design features, such as between or within subjects designs and repeated measures.

Experimental approaches can be used to test potential interventions. Protzko et al. (\protect\hyperlink{ref-protzko2020}{2020}) showed, across 16 studies, that better methodological practices led to replication rates that matched theoretical expectations with replication effect sizes comparable with the original. Many Labs 5 added expert (original author) advice to a replication process, and found that it did not substantially increase the replication rate for the 10 studies (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). These types of experiments are valuable for testing potential causes of non-replication, but they don't scale well due to expense, and not all influences on non-replication may be experimentally manipulable.

Given the limitations of experimental approaches, we have to rely on correlational approaches for now. Correlational approaches depend on data from replications, generally drawing heavily from the same small set of data points. In particular, the RP:P dataset itself is much discussed and reanalyzed (\protect\hyperlink{ref-anderson2016}{Anderson et al. 2016}, \protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}) to the point that much of what we think we know about replicability may be overfit to the 100 studies included in in RP:P. We need more data to find potential targets of concern and intervention.

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{question-summary-and-takehomes}{%
\subsection{question, summary, and takehomes}\label{question-summary-and-takehomes}}

NEED TO REDO!
MAY NEED SMOOTHER TRANSITIONS THROUGHOUT
Prior large scale replications of psychology findings have estimated the replication rate at somewhere around one half, which intuitively feels low. However, limited numbers of replications and limited research into specific predictors of replication failure mean that the reasons for non-replications are not well understood.

@@

The same issue that may cause a study not to replicate under constrained circumstances (despite hypothetically replicating under more favorable circumstances such as expert administration or larger budgets) will also plague attempts to build off those studies in constrained circumstances. Thus, we believe it is relevant to estimate replicability under the limitations of real world resource and expertise constraints.
\#\# intepretation - replication in light of the goal

Replication results should be interpreted in light of their methods and estimand.

Given our sample, we are able to estimate the correlates of studies replicating when the replications are conducted online, by graduate students with limited time and budget. We do not interpret our non-replications as indicating the original results were false positives (presumably some were and some weren't). There are many possible reasons for the non-replications in this sample. In some cases, the problem may be with the replication, such as too few participants, many exclusions for failed attention checks, or participants speeding through the study. When these issues are diagnosed, they suggest possible ways to ``rescue'' the replication by increasing the sample or changing the interface, without altering the underlying experiment; thus, while the replication did not succeed, after some troubleshooting, students may still be able to extend the work in the future. In other cases, there were a priori reasons to distrust the original study, such as exclusion criteria that seemed post-hoc or high-order interaction terms with a small sample. That said, not all scientists recognize the same factors as potential indications of low power or questionable research practices; students conducting these replications generally expected them to succeed. In many non-replication cases, it was unclear why the results failed to replicate.

\hypertarget{limitations}{%
\subsection{limitations}\label{limitations}}

We conducted an observational study of replication attempts, and our results are limited by the studies we included, which are limited in number and may not be representative of the studies of interest to psychologists as a whole. Our predictor variables were not manipulated, so they cannot be interpreted as causing (non-) replication, but only as correlational markers. Some of the correlates are most easily interpreted as being about the original study, and others reflect the closeness of the replication to the original. For instance, while within-subjects designs are more likely to replicate than between-subjects designs, this could be related to power, or the types of experiments that tend to be run in each design. Given the predictive value, slightly more skepticism and critical reading of between-subjects designs may be warranted, but this correlation, by itself, does not mean scientists should prefer to run within-subjects designs.

\hypertarget{conclusion}{%
\subsection{conclusion}\label{conclusion}}
